# loadflexBatch

This repository contains a demo script and data for running the `loadflex`
package in batch mode, i.e., for multiple sites, constituents, and models.

# Quick Start

1. Open a new R session.

2. Option A: If you've never installed `loadflex`, run these commands to install
`loadflex` and its dependencies.
```{r, eval=FALSE}
install.packages(
  c("devtools", "smwrData", "smwrBase", "smwrGraphs", "smwrStats", "smwrQW", "rloadest", "unitted"), 
  repos=c("https://owi.usgs.gov/R", "https://cran.rstudio.com"), 
  dependencies=TRUE, type="both")
install.packages(
  c("car", "dplyr", "ggplot2", "lubridate", "MASS", "Matrix"),
  dependencies=TRUE, type="both")
devtools::install_github("USGS-R/loadflex")
```

2. Option B: If you've installed `loadflex` in the past, run these commands to
get any updates to `loadflex` or its dependencies.
```{r, eval=FALSE}
update.packages(
  oldPkgs=c("smwrData", "smwrBase", "smwrGraphs", "smwrStats", "smwrQW", "rloadest", "unitted"),
  dependencies=TRUE, repos=c("https://owi.usgs.gov/R", "https://cran.rstudio.com"))
update.packages(
  oldPkgs=c("car", "dplyr", "ggplot2", "lubridate", "MASS", "Matrix"),
  dependencies=TRUE, type="both")
devtools::install_github("USGS-R/loadflex")
```

3. Move into the directory where you want to add loadflexBatch as a
subdirectory. For example, if you want to use your home directory:
```{r, eval=1, echo=1}
setwd("~")
tdir <- tempdir()
setwd(tdir)
```

4. Download and unzip the files in this batch mode demo.
```{r}
download.file("https://github.com/USGS-R/loadflexBatch/archive/master.zip", "loadflexBatch.zip")
unzip("loadflexBatch.zip")
success <- file.remove("loadflexBatch.zip")
```

5. Change into the loadflexBatch-master subdirectory and look around: you should
see files including this `README.md`, the `batch.R` script, and the example
input data folder `three_ANA_sites`.
```{r}
setwd("loadflexBatch-master")
dir()
```

6. Open `batch.R` to read through the script and to change options in the User
Inputs section if desired.
```{r, eval=FALSE}
file.edit('batch.R')
```

7. Run `batch.R`. You can do this from RStudio using the Source button at the
top right of the editor pane (or using Ctrl+Shift+S/Cmd+Shift+S), or you can run
the following line:
```{r, eval=FALSE}
source('batch.R')
```

8. Inspect the outputs. These are in the `output` directory (unless you've
changed the `outputFolder` setting in the User Inputs section of `batch.R`).
You'll probably want to open that directory in a file explorer, but you can also
explore the contents from within R:
```{r, echo=-1}
setwd("loadflexBatch-master")
dir('output')
head(read.csv('output/annual.csv', header=TRUE, stringsAsFactors=FALSE))
```

```{r, echo=FALSE}
# Clean up after ourselves
unlink("loadflexBatch-master", recursive=TRUE)
```
